---
title: "Attrition Analysis"
author: "Eric Cadena"
date: "7/29/2021"
output: html_document
Link to Youtube:  https://www.youtube.com/watch?v=up0r6qSmRMk
---

#Executive Summary
DDS Analytics has provided an employee data set and wants to know the key factors
leading to an employee leaving the Frito Lay company.  After conducting Explanatory
Data Analysis, Linear Regression models and creating a predictive model with a
greater than 60% in Accuracy, Specificity and Sensitivity, it was concluded that
being in a Manager role vs. a Non Manager role, Yearly Salary and Stock Option Levels
were the biggest predictors of Attrition.  If you are in a Non Manager role, earning under $50k
a year, and you are not invested in the Company through Stock Options, there is a strong chance
you will leave the Company.  An interesting secondary finding is that relationships matter to attrition in the following way; if you have been with the same manager for over 2 years, you are likely to stay with the company and if you are a divorced employee you are more likely to stay with the company.

#Introduction
The analysis was conducted using, heavy visualization for the purpose
of conducting a thorough Explanatory Data Analysis while looking at the
relationship between and among several variables.  Once hints emerged as to
the key factors leading to attrition, a linear regression model was implemented
and prediction tests were run until a predictive model was built that could
accurately tell us which factors can be identified an an employee that puts
him at risk of leaving the company.

```{r}
#Packages used
library(tidyverse)
library(ggplot2)
library(plotly)
library(class)
library(caret)
library(e1071)
library(plyr)
library(ROSE)
library(scales)
library(plotly)
library(data.table)

#Set up Primary Data Source in R
setwd('C:/Users/181429/Desktop/Semester 1 Courses/Doing Data Science/Unit 14 and 15 Case Study 2')  ## Set working directory
work <- read.csv("C:/Users/181429/Desktop/Semester 1 Courses/Doing Data Science/Unit 14 and 15 Case Study 2/CaseStudy2-data.csv", stringsAsFactors = TRUE, na.strings='..')
work = data.frame(work)
```

```{r}
#PART 1 OF THE ANALYSIS IS DATA TIDYING 
#Change JobRole Cells to be more readable in a visual display
work = work%>%mutate(JobRole=as.character(JobRole))%>%mutate(JobRole = replace(JobRole, JobRole=='Sales Executive', 'Sale Exe'))%>%
  mutate(JobRole = replace(JobRole, JobRole=='Research Director', 'Res Dir'))%>%
  mutate(JobRole = replace(JobRole, JobRole=='Manufacturing Director', 'Man Dir'))%>%
  mutate(JobRole = replace(JobRole, JobRole=='Research Scientist', 'Res Sci'))%>%
  mutate(JobRole = replace(JobRole, JobRole=='Sales Representative', 'Sale Rep'))%>%
  mutate(JobRole = replace(JobRole, JobRole=='Healthcare Representative', 'Health Rep'))%>%
  mutate(JobRole = replace(JobRole, JobRole=='Manager', 'MGR'))%>%
  mutate(JobRole = replace(JobRole, JobRole=='Human Resources', 'HR'))%>%
  mutate(JobRole = replace(JobRole, JobRole=='Laboratory Technician', 'Lab Tech'))


#Create a yearly Salary column
#Creating Salary
work$Salary = work$MonthlyIncome*12

#Manager vs Not Manager
work1 <- within(work, MGR <- ifelse(JobRole=="Sale Exe","MANAGER",
ifelse(JobRole=="Lab Tech","NOT MANAGER",
ifelse(JobRole=="HR", "MANAGER",
ifelse(JobRole=="MGR","MANAGER",
ifelse(JobRole=="Health Rep", "NOT MANAGER",
ifelse(JobRole=="Sale Rep", "NOT MANAGER",
ifelse(JobRole=="Res Sci", "NOT MANAGER",
ifelse(JobRole=="Man Dir", "MANAGER",
ifelse(JobRole=="Res Dir", "MANAGER",NA))))))))))

#Creating Columns to create binary codes and categorize (Manager-1 vs Not Manager-0), (Single/Married-0 vs Divorced-1),
#(Attrition-0 vs Not-Attrition-1)
work2 <- within(work1, MGRcode<- ifelse(MGR=="MANAGER","1", ifelse(MGR=="NOT MANAGER","0",NA)))

work3 <- within(work2, Maritalcode<- ifelse(MaritalStatus=="Single","0",
ifelse(MaritalStatus=="Married","0",
ifelse(MaritalStatus=="Divorced","1", NA))))

work4 <- within(work3, Attritioncode<- ifelse(Attrition=="No","1", ifelse(Attrition=="Yes","0",NA)))

work5 <- within(work4, Stockcode <- ifelse(StockOptionLevel=="0","Not Invested",
ifelse(StockOptionLevel=="1","Sort of Invested",
ifelse(StockOptionLevel=="2","Invested",
ifelse(StockOptionLevel=="3","Really Invested",NA)))))

work6 <- within(work5, Timemgr <- ifelse(YearsWithCurrManager=="0","No Relationship",
ifelse(YearsWithCurrManager=="1","No Relationship",
ifelse(YearsWithCurrManager=="2","Some Relationship",
ifelse(YearsWithCurrManager=="3","Some Relationship",
ifelse(YearsWithCurrManager=="4","Some Relationship",
ifelse(YearsWithCurrManager=="5","Personal Relationship",
ifelse(YearsWithCurrManager=="6","Personal Relationship", 
ifelse(YearsWithCurrManager=="7","Personal Relationship",
ifelse(YearsWithCurrManager=="8","Personal Relationship", 
ifelse(YearsWithCurrManager=="9","Personal Relationship",
ifelse(YearsWithCurrManager=="10","Personal Relationship", 
ifelse(YearsWithCurrManager=="11","Personal Relationship",
ifelse(YearsWithCurrManager=="12","Personal Relationship", 
ifelse(YearsWithCurrManager=="13","Personal Relationship",
ifelse(YearsWithCurrManager=="14","Personal Relationship", 
ifelse(YearsWithCurrManager=="15","Personal Relationship", 
ifelse(YearsWithCurrManager=="16","Persoanl Relationship",
ifelse(YearsWithCurrManager=="17","Personal Relationship",NA)))))))))))))))))))

#Creating Salary Ranges
work7 <- within(work6, Range <- ifelse(Salary<50000, "0k-50k",
ifelse(Salary>50000 & Salary < 75000, "50k-75k",
ifelse(Salary>75000 & Salary < 100000, "75k-100k",
ifelse(Salary > 100000, ">100k", NA))))) 


#Coding Salaries to conduct Statistics tests
work8 = within(work7, Range2 <- ifelse(Salary<=20000, "20", 
ifelse(Salary>20000 & Salary<=25000, "25", 
ifelse(Salary>25000 & Salary<=30000, "30",
ifelse(Salary>30000 & Salary<=35000, "35",
ifelse(Salary>35000 & Salary<=40000, "40",
ifelse(Salary>40000 & Salary<=45000, "45",
ifelse(Salary>45000 & Salary<=50000, "50",
ifelse(Salary>50000 & Salary<=55000, "55",
ifelse(Salary>55000 & Salary<=60000, "60", 
ifelse(Salary>60000 & Salary<=65000, "65",
ifelse(Salary>65000 & Salary<=70000, "70",
ifelse(Salary>70000 & Salary<=75000, "75",
ifelse(Salary>75000, "80",NA))))))))))))))

#Coding Business Travel
work9 = within(work8, Travel <- ifelse(BusinessTravel=="Travel_Rarely","1",
ifelse(BusinessTravel=="Travel_Frequently","0",
ifelse(BusinessTravel=="Non-Travel","2",NA))))
      
#WE ARE NOT CONCERNED WITH THOSE HIRING EMPLOYEES and so we will Eliminate HR from the analysis.
##WE ARE NOT CONCERNED WITH THE TOP OF THE COMPANY HEIRARCHY 
#On a side note, HR is an insignificant portion of those leaving the company
work10 = filter(work9, JobRole != "HR")
work10
```

```{r}
#CONDUCT AND EXPLORATORY DATA ANALYSIS (EDA)
#Attrition Rate
ggplot(work10, aes(x=Attrition))+
  geom_bar(aes(y=(..count..)/sum(..count..)))+
  geom_text(aes(y=((..count..)/sum(..count..)),
        label=scales::percent((..count..)/sum(..count..))),
        stat="count",vjust=-0.25)


#Attrition MGR vs Not Manager
ggplot(work10, aes(x=MGR, fill=Attrition))+
  geom_bar(aes(y=(..count..)/sum(..count..)))+
  geom_text(aes(y=((..count..)/sum(..count..)),
                label=scales::percent((..count..)/sum(..count..))),
            stat="count",vjust=-0.25)
                                  
#Salary and Attrition
ggplot(work10, aes(x=as.character(Range), fill=Attrition))+
  geom_bar(aes(y=(..count..)/sum(..count..)))+
  geom_text(aes(y=((..count..)/sum(..count..)),
                label=scales::percent((..count..)/sum(..count..))),
            stat="count",vjust=-0.25)

#Job Role and Attrition
ggplot(work10, aes(x=JobRole, fill=Attrition))+
  geom_bar(aes(y=(..count..)/sum(..count..)))+
  geom_text(aes(y=((..count..)/sum(..count..)),
                label=scales::percent((..count..)/sum(..count..))),
            stat="count",vjust=-0.25)


#EDA showing relationship between Earnings and Job Role with a look at a variety of variables like Number of Companies Worked, Education, Stock Option Level, Years at Company,
#Years in Current Role, overtime, Sex, Marital Status, Business Travel & Department 
ggplot(data=work10)+geom_bar(mapping=aes(x=JobRole, fill=Attrition))
ggplot(data=work10)+geom_point(mapping=aes(x=NumCompaniesWorked, y=Salary, color=Attrition, size=Attrition, shape=Attrition))+facet_grid(JobRole~OverTime)
ggplot(data=work10)+geom_point(mapping=aes(x=Education, y=Salary, color=Attrition, size=Attrition, shape=Attrition))+facet_grid(JobRole~Gender)
ggplot(data=work10)+geom_point(mapping=aes(x=StockOptionLevel, y=Salary, color=Attrition, size=Attrition, shape=Attrition))+facet_grid(JobRole~MaritalStatus)
ggplot(data=work10)+geom_point(mapping=aes(x=YearsAtCompany, y=Salary, color=Attrition, size=Attrition, shape=Attrition))+facet_grid(JobRole~BusinessTravel)
ggplot(data=work10)+geom_point(mapping=aes(x=YearsInCurrentRole, y=Salary, color=Attrition, size=Attrition, shape=Attrition))+facet_grid(JobRole~Department)



#The data shows a clear relationship between the Job Role that leads to Attrition?


#EDA ON STAYING OR LEAVING IF YOU ARE IN A HIGH ATTRITION JOB BASED ON A VARIABLE OTHER THAN SALARY
#with current manager (*Postiive)
ggplot(work10, aes(x=factor(Attrition)))+geom_bar(aes(y=(..count..)/sum(..count..), fill=as.factor(Timemgr)))+
  scale_y_continuous(labels=percent)+xlab("No=Stay with Co. Yes=Leave Co.")+ylab("Overall Percent")+ggtitle("0 Isolates 0-1 year with Curr MGr")

#with stock options (*Positive)
ggplot(work10, aes(x=factor(Attrition)))+geom_bar(aes(y=(..count..)/sum(..count..), fill=as.factor(Stockcode)))+
  scale_y_continuous(labels=percent)+xlab("No=Stay with Co. Yes=Leave Co")+ylab("Overall Percent")+ggtitle("0 Isolates 0 or 1 Level Stock Option")

#marital Status (*Positive)
ggplot(work10, aes(x=factor(Attrition)))+geom_bar(aes(y=(..count..)/sum(..count..), fill=as.factor(Maritalcode)))+
  scale_y_continuous(labels=percent)+xlab("No=Stay with Co. Yes=Leave Co")+ylab("Overall Percent")+ggtitle("0 Isolates Single/Married")

#with work life balance (*Even employees happy with this are leaving)
ggplot(work10, aes(x=factor(Attrition)))+geom_bar(aes(y=(..count..)/sum(..count..), fill=as.factor(WorkLifeBalance)))+
  scale_y_continuous(labels=percent)+xlab("No=Stay with Co. Yes=Leave Co")+ylab("Overall Percent")+ggtitle("Work Life Balance at 4 levels")

#Relationship Satisfaction (*Not a Factor)
ggplot(work10, aes(x=factor(Attrition)))+geom_bar(aes(y=(..count..)/sum(..count..), fill=as.factor(RelationshipSatisfaction)))+
  scale_y_continuous(labels=percent)+xlab("No=Stay with Co. Yes=Leave Co")+ylab("Overall Percent")+ggtitle("Relationship Satisfaction at 4 levels")

#Environment Satisfaction (*Even employees happy with this are leaving)
ggplot(work10, aes(x=factor(Attrition)))+geom_bar(aes(y=(..count..)/sum(..count..), fill=as.factor(EnvironmentSatisfaction)))+
  scale_y_continuous(labels=percent)+xlab("No=Stay with Co. Yes=Leave Co")+ylab("Overall Percent")+ggtitle("Environment Satisfaction at 4 levels")

#Job Satisfaction  (*Not a Factor)
ggplot(work10, aes(x=factor(Attrition)))+geom_bar(aes(y=(..count..)/sum(..count..), fill=as.factor(JobSatisfaction)))+
  scale_y_continuous(labels=percent)+xlab("No=Stay with Co. Yes=Leave Co")+ylab("Overall Percent")+ggtitle("Job Satisfaction at 4 levels")

#Business Travel (*Not a Factor)
ggplot(work10, aes(x=factor(Attrition)))+geom_bar(aes(y=(..count..)/sum(..count..), fill=as.factor(BusinessTravel)))+
  scale_y_continuous(labels=percent)+xlab("No=Stay with Co. Yes=Leave Co")+ylab("Overall Percent")+ggtitle("Business Travel at 3 Categories")
```

```{r}
#WE ARE READY TO CREATE A MODEL USING FOUR VARIABLES DISCOVERED IN THE EDA THAT WE THINK IMPACTS THE DECISION TO LEAVE THE COMPANY
#Isolating the columns we want to use to build the model
Attrition = select(work10, MGR,MaritalStatus, Stockcode,Range,BusinessTravel, Timemgr, Attrition)
Attrition
summary(Attrition)

#Data Partition
set.seed(123)
ind = sample(seq(1:length(Attrition$Attrition)),round(.7*length(Attrition$Attrition)))
train = Attrition[ind,]
test = Attrition[-ind,]

model = naiveBayes(Attrition ~ MGR + MaritalStatus + Stockcode + Range + Timemgr, data=train)
confusionMatrix(predict(model,test), test$Attrition)

#Over Sample
over <- ovun.sample(Attrition~.,data=train, method="over",N=998)$data
table(over$Attrition)

model2 = naiveBayes(Attrition ~ MGR + MaritalStatus + Stockcode + Range, data = over)
confusionMatrix(predict(model2,test), test$Attrition)


#Saving csv prediction file for Attrition
A = predict(model2,test)
A
A1=data.frame(A)

setDT(A1, keep.rownames=TRUE)[]
A2 = setNames(A1, c("ID","Attrition"))
A2

write.csv(A2, file="C:/Users/181429/Desktop/Semester 1 Courses/Doing Data Science/Unit 14 and 15 Case Study 2//Case2PredicitonsCadenaAttrition.csv", row.names=FALSE)
```

```{r}
#WHAT IS AN EMPLOYEE MOST LIKELY TO MAKE
#Find a relationship between the response variable Income and an explanatory variable
work9%>%ggplot(aes(x=TotalWorkingYears, y=MonthlyIncome))+geom_point()+ggtitle("Relationship between Working Years and Monthly Rate")+geom_smooth(method="lm")

work11 = select(work9,ID, MonthlyIncome, TotalWorkingYears, JobLevel)


#Multiple Linear Regression
set.seed(123)
ind = sample(seq(1:length(work11$MonthlyIncome)),round(.7*length(work11$MonthlyIncome)))
train = work11[ind,]
test = work11[-ind,]

M = lm(MonthlyIncome~TotalWorkingYears+factor(JobLevel), data=train)
M
par(mfrow=c(2,2))
plot(M)

p2 = predict(M,test)
error = test$MonthlyIncome-p2
RMSE = sqrt(mean(error^2))
RMSE
p2

#Saving csv prediction file for Monthly Income
p3=data.frame(p2)
p3

setDT(p3, keep.rownames=TRUE)[]
p4 = setNames(p3, c("ID","MonthlyIncome"))
p4

write.csv(p4, file="C:/Users/181429/Desktop/Semester 1 Courses/Doing Data Science/Unit 14 and 15 Case Study 2//Case2PredicitonsCadenaSalary.csv", row.names=FALSE)
```





